---
title: "MIBI cell clustering"
author: "SJ & YYY"
date: "06/23/2022"
output: html_document
---

## R Markdown
This markdown is to generate cell clustering maps from cellular features extracted from CODEX.
Before starting, install these packages if necessary:

BiocManager::install(c("FlowSOM", "MetaCyto", "Biobase", "EBImage", "limma"))
install.packages(c("matrixStats", "ggplot2", "reshape2", "dplyr", "ggrepel", "RColorBrewer", "pheatmap", "ConsensusClusterPlus", "tidyverse", "Rtsne", "uwot", "ComplexHeatmap", "bioimagetools", "magick"))
install.packages("remotes")
remotes::install_github("JonathanIrish/MEMv3")
remotes::install_github("jmw86069/jamba")

```{r, echo=FALSE, warning=FALSE, error=FALSE}
#Import these libraries. They are needed to operate cell clustering.
#Remember to change the "source" file directory to where the MIBI_functions.R is located
library(flowCore)
library(R.matlab)
library(FlowSOM)
library(MetaCyto)
library(matrixStats)
library(ggplot2)
library(reshape2)
library(dplyr)
library(limma)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
library(ConsensusClusterPlus)
library(Biobase)
library(MEM)
library(tidyverse)
library(Rtsne)
library(uwot)
library(ComplexHeatmap)
library(bioimagetools)
library(magick)
library(EBImage)
library(jamba)
library(patchwork)
source("/       [directory]        /MIBI_functions.R") # please update directory 
```

Import .fcs file after cell feature extraction
```{r}
# Set file directory
directory = "/Users/Desktop/dataScaleSize.fcs"

# Import file
fcs_no_trans = read.flowSet(files = directory, transformation = F, truncate_max_range = F)
fcs_no_trans
```

Make sure that every cell has nucleus.
```{r}
# exprs retrieves expression data and returns a large matrix of expression values
expr_notrans = fsApply(fcs_no_trans, exprs)

# Check that every cell has nuclear signal. If the value returns "0", it means that there are no cells with no nuclear signal.
sum(expr_notrans[, "Histone H3"] == 0)
```

Rename channel names if necessary
```{r}
# set file directory for channel names
channel_names = read.csv('/Users/channelNames_analysis.csv')

# actual data column starts at column #3
for (col in 3:ncol(expr_notrans)){
  colnames(expr_notrans)[col] <- channel_names[col-2,1] # the first 2 columns are cellLabel and cellArea
  }

# append "PointNum" column
PointNum = rep(c(1), each=nrow(expr_notrans))
expr_notrans = cbind(expr_notrans, PointNum)

colnames(expr_notrans)
```

Set nuclear normalization parameters.
```{r}
# Set normalization channel and cutoff signal; the prevailing assumption is that norm_channel signal should be generally consistent across diff FOVs
norm_channel = "Histone H3"
cutoff = 0.1

# Exclude cells that have nucleus below the cutoff
expr_notrans_filter = expr_notrans[expr_notrans[, norm_channel] > cutoff, , drop=F]

# Check the number of cells in the original vs. the filtered file.
dim(expr_notrans)
dim(expr_notrans_filter)
```

Normalize all channels to nucleus by dividing each signal with the median nuclear signal
```{r}
# Perform normalization to the median nuclear signal (exclude cellLabelInImage, cellSize, PointNum)

# Exclude "PointNum" column
normChannels = ncol(expr_notrans)-1

# Normalize for each PointNum. Usually, each "PointNum" refers to a new imaging dataset, but here the "PointNum" = 1
expr_notrans_filter_medianNucleus = list()
for(i in 1:max(expr_notrans_filter[,"PointNum"])){
  w = which(expr_notrans_filter[,"PointNum"] == i)
  df = expr_notrans_filter[w,]
  df[,3:normChannels] = df[,3:normChannels]/(median(df[,norm_channel])) # exclude cellLabelInImage and cellSize columns
  expr_notrans_filter_medianNucleus[[i]] = df
  }

expr_notrans_filter_medianNucleus = do.call(
  rbind, lapply(expr_notrans_filter_medianNucleus, function(x){return(x)})
  )

# Show number of cells after normalization, as well as the before/after normalization values
head(expr_notrans_filter)
head(expr_notrans_filter_medianNucleus)
```

Perform arcsinh data transformation and percentile cutoff + scaling (from 0 - 1)
```{r}
# Adjust "cofactor" to change arcsinh transformation. 
cofactor = 5

# Perform arcsine transformation
expr_notrans_filter_medianNucleus[,3:normChannels] = asinh(expr_notrans_filter_medianNucleus[,3:normChannels]/cofactor)
expr_notrans_filter_medianNucleus_original = expr_notrans_filter_medianNucleus

# Set percentile (two-sided cutoff)
percentile = 99

# Determine percentile range
percentileBot = 1-percentile/100
percentileTop = percentile/100

range = colQuantiles(
  expr_notrans_filter_medianNucleus[, 3:normChannels], # exclude cellLabelinImage, cellSize, PointNum
  probs = c(percentileBot, percentileTop))

# Scale the data accordingly
expr = t((t(expr_notrans_filter_medianNucleus[, 3:normChannels])-range[,1]) / (range[,2]-range[,1]))
expr[expr<0] = 0
expr[expr>1] = 1

# Add back cellLabelinImage, cellSize, and PointNum columns
expr_notrans_filter_medianNucleus[,3:normChannels] = expr

head(expr_notrans_filter_medianNucleus_original)
head(expr_notrans_filter_medianNucleus)
```

Plot graphs for unscaled and scaled data. A "tip" at the x = 1 is indicative of positive signals
```{r, fig.width=9, fig.height=9}
# Unscaled data
ggdf_original = data.frame(expr_notrans_filter_medianNucleus_original)
ggdf_original = melt(ggdf_original, id.var = "PointNum",
                     value.name = "expression",
                     variable.name = "antigen")
ggdf_original$PointNum = as.factor(ggdf_original$PointNum)
g_original = ggplot(ggdf_original, aes(x = expression, color = PointNum, group = PointNum))

# Custom palette for unscaled data graph
getPalette = colorRampPalette(brewer.pal(9, "Set1"))
g_original + geom_density() + facet_wrap(~ antigen, nrow = 10, scales = "free") + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 10), 
        axis.text = element_text(size = 10)) + 
  scale_fill_manual(values=getPalette(length(unique(ggdf_original$PointNum))))


# Scaled data
ggdf = data.frame(expr_notrans_filter_medianNucleus)
ggdf = melt(ggdf, id.var = "PointNum",
            value.name = "expression",
            variable.name = "antigen")
ggdf$PointNum = as.factor(ggdf$PointNum)
g = ggplot(ggdf, aes(x = expression, color = PointNum, group = PointNum))

# Custom palette for scaled data graph
getPalette = colorRampPalette(brewer.pal(9, "Set1"))
g + geom_density() + facet_wrap(~ antigen, nrow = 10, scales = "free") + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 10), 
        axis.text = element_text(size = 10)) + 
  scale_fill_manual(values=getPalette(length(unique(ggdf$PointNum))))
```

Select markers to be used for UMAP. USE ONLY SELECTED MARKERS FOR PHENOTYPING, NOT ALL MARKERS!
```{r}
# Select the markers; do not include cellLabelInImage, cellSize, nucleus, PointNum.
# Use colnames(expr_adjusted_df) to facilitate the search for columns.

# colnames(expr_adjusted_df)
umap.markers = expr_adjusted_df[, c("marker1","marker2","marker3","...")]

colnames(umap.markers)
umap.markers
```

Perform UMAP and generate UMAP plot
```{r run_UMAP, fig.height=8, fig.width=10}
## Getting colors here
values = c(brewer.pal(n = 9, name = "Set1"), brewer.pal(n = 12, name = "Paired"), brewer.pal(n = 8, name = "Set2"),
           brewer.pal(n = 8, name = "Dark2"), brewer.pal(n = 12, name = "Set3"), brewer.pal(n = 8, name = "Accent") )

# Run UMAP on the chosen markers above
set.seed(1)
myumap <- umap(umap.markers, 
               ret_model = TRUE, 
               verbose = TRUE)
umap.data = as.data.frame(myumap$embedding)   # pull out umap values from 'myumap' into a dataframe
colnames(umap.data) <- c("UMAP1", "UMAP2")    # assign column names to umap.data
range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio <- (range[1]/range[2])

# Generate UMAP flat dotplot and density dotplot (1 dot = 1 cell)
UMAP.plot <- data.frame(x = umap.data[,1], y = umap.data[,2])
## flat dotplot
ggplot(UMAP.plot) + 
  coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y), cex = 1) + 
  labs(x = "UMAP 1", y = "UMAP 2") + theme_bw()
## density dotplot
ggplot(UMAP.plot, aes(x=x, y=y)) + 
  coord_fixed(ratio = graphical.ratio) + 
  geom_bin2d(bins = 128) + 
  scale_fill_viridis_c(option = "A", trans = "sqrt") + 
  scale_x_continuous(expand = c(0.1,0)) + 
  scale_y_continuous(expand = c(0.1,0)) + 
  labs(x = "UMAP 1", y = "UMAP 2") + theme_bw()

# Color by FOV
PointNum = factor(expr_adjusted_df[,"PointNum"], 
                  levels = unique(expr_adjusted_df[,"PointNum"])[
                    mixedOrder(unique(expr_adjusted_df[,"PointNum"]))])

# Plot UMAP
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=PointNum),cex = 0.5) + 
  guides(colour = guide_legend(override.aes = list(size=5), nrow = 13)) +
  labs(x = "UMAP 1", y = "UMAP 2" ,title = "FlowSOM Clustering on UMAP Axes", color = "FOV") + 
  theme_bw() + scale_color_manual(values = values)
```

UMAP plotting for each marker
```{r, fig.height=5, fig.width=10}
# Generate complete data: filtered data + umap data
full_umap = cbind(umap.markers, umap.data)

# Make list of plots
channel_to_plot = colnames(full_umap)
## unscaled
per_channel_list1 = lapply(channel_to_plot, function(channel){
  print(channel)
  plot_gradient_mean(full_umap, channel, midscale = F)
  })
names(per_channel_list1) = channel_to_plot
## scaled
per_channel_list2 = lapply(channel_to_plot, function(channel){
  plot_gradient_mean(full_umap, channel, midscale = T)
  })
names(per_channel_list2) = channel_to_plot

# Assemble with wrap_plots()
require(patchwork) # require is generally used inside functions
wrap_plots(per_channel_list1, ncol = 5)
wrap_plots(per_channel_list2, ncol = 5)
```

Perform FlowSOM clustering based on the UMAP axes; therefore, one should have already done UMAP based on the selected markers
```{r run_FlowSOM_on_UMAP, fig.height=8, fig.width=8}
# Enter target number of clusters
target.clusters = 50

# Running flowsom on the actual selected markers, instead of the UMAP reduced dimensions
flowsom_markers = umap.markers
colnames(flowsom_markers)
mat <- as.matrix(flowsom_markers)

# Create flowFrame for FlowSOM input
metadata <- data.frame(name = dimnames(mat)[[2]],
                       desc = dimnames(mat)[[2]])
metadata$range <- apply(apply(mat, 2, range), 2, diff)
metadata$minRange <- apply(mat, 2, min)
metadata$maxRange <- apply(mat, 2, max)
input.flowframe <- new("flowFrame",
                       exprs = mat,
                       parameters = AnnotatedDataFrame(metadata))

# Implement FlowSOM on the data
fsom <- FlowSOM(input.flowframe,
                colsToUse = c(1:dim(input.flowframe)[2]),
                nClus = target.clusters,
                seed = 1
                )

# This gets the cluster information for each cell
FlowSOM.clusters <- GetMetaclusters(fsom)
#FlowSOM.clusters <- as.matrix(fsom[[2]][fsom[[1]]$map$mapping[, 1]])

# Color palette
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, 
                           rownames(qual_col_pals)))
col_vector = col_vector[-c(4,17,19,27,29:45)]
values = sample(col_vector)

# Merge FlowSOM clusters on UMAP axes
umap.cent = cbind(UMAP.plot, FlowSOM.clusters) %>% 
  group_by(FlowSOM.clusters) %>% 
  summarize_all(mean)

# This allows the cluster to be plotted in numeric order
FlowSOM.clusters <- factor(FlowSOM.clusters, 
                           levels = unique(FlowSOM.clusters)[mixedOrder(unique(FlowSOM.clusters))])

# Plot flowsom clusters
##plot a figure with cluster names outside
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters), cex = 1.5) + 
  guides(colour = guide_legend(override.aes = list(size=5), nrow = 13)) +
  labs(x = "UMAP 1", y = "UMAP 2", 
       title = "FlowSOM Clustering on UMAP Axes", 
       color = "FlowSOM Cluster") + 
  theme_bw() + scale_color_manual(values = values)
##plot a figure with cluster names embedded in the figure
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters), cex = 1.5) + 
  guides(colour = guide_legend(override.aes = list(size=5), nrow = 13)) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "FlowSOM Clustering on UMAP Axes", 
       color = "FlowSOM Cluster") + 
  theme_bw() + scale_color_manual(values = values) + 
  geom_label_repel(data = umap.cent, aes(x = x, y = y, label = FlowSOM.clusters)) + 
  theme(legend.position = "none")
```

Marker Enrichment Modeling: creates quantitative identity labels based on features enriched in a population (relative to a reference population). 
```{r MEM, fig.height=5, fig.width=5}
# Run MEM on the FlowSOM clusters dervied from UMAP axes
cluster = as.numeric(as.vector((FlowSOM.clusters)))
MEM.data = cbind(flowsom_markers, cluster)
MEM.data_df = data.frame(MEM.data)

# Scale zero = off (zero.ref = FALSE)
MEM.values.uf_zeroOFF = MEM(
  MEM.data_df,
  transform = FALSE,
  cofactor = 0,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "none",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
  )

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf_zeroOFF,
  cluster.MEM = "both",
  cluster.medians = "none",
  cluster.IQRs = "none",
  display.thresh = 1,
  newWindow.heatmaps = F,
  output.files = T,
  labels = T,
  only.MEMheatmap = T
  )

# Scale zero = on (zero.ref = TRUE)
MEM.values.uf_zeroON = MEM(
  MEM.data_df,
  transform = FALSE,
  cofactor = 0,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = TRUE,
  rename.markers = FALSE,
  new.marker.names = "none",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
  )

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf_zeroON,
  cluster.MEM = "both",
  cluster.medians = "none",
  cluster.IQRs = "none",
  display.thresh = 1,
  newWindow.heatmaps = F,
  output.files = T,
  labels = T,
  only.MEMheatmap = T
  )

MEM.data_df
```

Export cluster values; here, visually check and annotate each cell
```{r}
# Select table
cell_cluster = cbind(expr_adjusted_df$Ch.1,MEM.data_df$cluster)
colnames(cell_cluster) = c("cellID", "cluster")
head(cell_cluster)

# Write cluster
write.csv(cell_cluster, "/Users/Desktop/MIBIclusters.csv")
```

